---
title: '`tidyverse`: First Encounter'
date: "`r format(Sys.time(), '%d %B %Y')`" 
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:
      ctex: UTF8
    number_sections: yes
classoptions: hyperref, 12pt, a4paper
---

# `tidyverse`: First Encounter

## Prerequisites

### loading `tidyverse` 
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

### loading dataset `nycflights13`
```{r, message=FALSE, warning=FALSE}
library(nycflights13)
```

### print the sub dataset `flights` from `nycflights13`
```{r}
flights
```

## Choosing columns: `select`, `rename`

### pick columns: `select()` 
```{r}
flights %>% select(carrier, flight)
```

### hide columns: `select()` with minus signs
```{r}
flights %>% select(-month, -day)
```

### hide a range of columns: range
```{r results='hide'}
flights %>% select(-(dep_time:arr_delay))
```

### hide any column with a matching name: `contains()`
```{r results='hide'}
flights %>% select(-contains("time"))
```

### pick columns using a character vector of column names
```{r}
cols <- c("carrier", "flight", "tailnum")
flights %>% select(all_of(cols))
```

### `all_of()` vs `any_of()`
```{r}
cols <- c("carrier", "flight", "tailnum_ttt")
flights %>% select(any_of(cols))
```

### rename columns with that all columns not mentioned are dropped: `select()`
```{r}
flights %>% select(tail = tailnum)
```

### rename columns with that all columns not mentioned are kept: `rename()`
```{r}
flights %>% rename(tail = tailnum)
```

## Choosing rows: `filter`, `between`, `slice`, `slice_sample`, `slice_max`, `distinct`

### `filter()` supports the use of multiple conditions
```{r}
flights %>% filter(dep_time >= 600, dep_time <= 605)
```

### if numeric values fall in a range: use `between()` 
```{r results='hide'}
flights %>% filter(between(dep_time, 600, 605))
```

* side note: `is.na()` is useful when filtering
```{ r results='hide'}
flights %>% filter(!is.na(dep_time))
```

### filter rows by position: `slice()` 
```{r}
flights %>% slice(1000:1005)
```

### keep the first five rows within each group: `group_by()` then `slice(1:5)` 
```{r}
flights %>%
    group_by(month, day) %>%
    slice(1:5)
```

### sample three rows from each group: `group_by()` then `slice_sample(n = 3)` 
```{r}
flights %>%
    group_by(month, day) %>%
    slice_sample(n = 3)
```

### keep three rows from each group with the top `dep_delay`: `group_by()` then `slice_max(dep_delay, n = 3)`
```{r}
flights %>%
    group_by(month, day) %>%
    slice_max(dep_delay, n = 3)
```

### sort by `dep_delay` within each group: `group_by()`, `slice_max()`, then `arrange(desc(dep_delay))`
```{r}
flights %>%
    group_by(month, day) %>%
    slice_max(dep_delay, n = 3) %>%
    arrange(desc(dep_delay))
```

### unique rows can be identified using `unique()` from base R
```{r}
flights %>%
    select(origin, dest) %>%
    unique()
```

### `tidyverse` provides an alternative that is more efficient: `distinct`
```{r results='hide'}
flights %>%
    select(origin, dest) %>%
    distinct()
```

* side note: when chaining, you don't have to include the parentheses if there are no arguments
```{r results='hide'}
flights %>%
    select(origin, dest) %>%
    distinct
```

## Adding new variables: `mutate`, `transmute`, `rownames_to_column`

### create new variables and keep all existing variables: `mutate()` 
```{r}
flights %>% mutate(speed = distance / air_time * 60)
```

### create new variables only: `transmute()` 
```{r}
flights %>% transmute(speed = distance / air_time * 60)
```

### `rownames_to_column()` turns row names into an explicit variable
```{r}
mtcars %>%
    rownames_to_column("model") %>%
    head()

# `tidyverse` does not print row names for local data frames
mtcars %>% as_tibble()
```

## Grouping and counting: `summarise`, `tally`, `count`, `group_size`, `n_groups`, `ungroup`

### count the number of rows in each group: `group_by()` then `summarize(cnt = n())`
```{r}
flights %>%
    group_by(month) %>%
    summarise(cnt = n())
```

### count the number of rows in each group: `group_by()` then `tally()`
```{r}
flights %>%
    group_by(month) %>%
    tally()
```

## Count the number of rows in each group: `count`
```{r}
flights %>% count(month)
```

### you can sort by the count: `arrange(desc(cnt))`
```{r}
flights %>%
    group_by(month) %>%
    summarise(cnt = n()) %>%
    arrange(desc(cnt))
```

### `tally()` and `count()` have a sort parameter for this purpose
```{r results='hide'}
flights %>%
    group_by(month) %>%
    tally(sort = TRUE)

flights %>% count(month, sort = TRUE)
```

### you can sum over a specific variable instead of simply counting rows
```{r}
flights %>%
    group_by(month) %>%
    summarize(dist = sum(distance))
```

### `tally()` and `count()` have a wt parameter for this purpose
```{r results='hide'}
flights %>%
    group_by(month) %>%
    tally(wt = distance)

flights %>% count(month, wt = distance)
```

### return the counts as a vector: `group_size()` 
```{r}
flights %>%
    group_by(month) %>%
    group_size()
```

### report the number of groups: `n_groups()` 
```{r}
flights %>%
    group_by(month) %>%
    n_groups()
```

### group by two variables, `summarise`, `arrange` (output is possibly confusing)
```{r}
flights %>%
    group_by(month, day) %>%
    summarise(cnt = n()) %>%
    arrange(desc(cnt)) %>%
    print(n = 10)
```

### arrange across all groups: `ungroup()` before arranging 
```{r}
flights %>%
    group_by(month, day) %>%
    summarise(cnt = n()) %>%
    ungroup() %>%
    arrange(desc(cnt))
```

## Creating data frames: `tibble`

`tibble()` is a better way than `data.frame()` for creating data frames. Benefits of `tibble()`:

* You can use previously defined columns to compute new columns
* It never coerces column types
* It never munges column names
* It never adds row names 
* It only recycles length 1 input
* It returns a local data frame

### `tibble()` example
```{r}
tibble(a = 1:6, b = a * 2, c = "string", "d+e" = 1) %>% glimpse()
```
### `data.frame()` example
```{r}
data.frame(a = 1:6, c = "string", "d+e" = 1) %>% glimpse()
```

## Viewing more output: `print`, `View`

### specify that you want to see more rows
```{r}
flights %>% print(n = 15)
```

### specify that you want to see ALL rows (don't run this!)
```{r eval=FALSE}
#flights %>% print(n = Inf)
```

### specify that you want to see all columns
```{r}
flights %>% print(width = Inf)
```

### show up to 1000 rows and all columns
```{r eval=FALSE}
flights %>% View

# set option to see all columns and fewer rows
#options(dplyr.width = Inf, dplyr.print_min = 6)

# reset options (or just close R)
#options(dplyr.width = NULL, dplyr.print_min = 10)
```

## Plotting: `ggplot2`

```{r}
flights %>%
    group_by(dest) %>%
    summarize(count = n(),
              dist = mean(distance, na.rm = TRUE),
              delay = mean(arr_delay, na.rm = TRUE)) %>%
    filter(delay > 0, count > 20, dest != "HNL") %>%
    ggplot(mapping = aes(x = dist, y = delay)) +
    geom_point(aes(size = count), alpha = 1 / 3) +
    geom_smooth(se = FALSE)
```
